---
title: "Introduction to R"
author: "Rebecca Barter"
date: "11/12/2021"
output: html_document
---


In this tutorial, you will learn the basics of R Markdown, data frames (tibbles), piping, data manipulations (select, filter, mutate, etc), and if there's time, visualization using ggplot2. Much of this is nested under the "tidyverse" R package suite.


## Data as data frames

First, let's load in the "Gapminder" data from the web:

```{r}
gapminder <- read.csv(
    "https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv"
)
```


We can look at the **first 6 rows** using the `head()` function:

```{r}

```


And the **dimension** of the dataset using `dim()`

```{r}

```

The data contains 6 "columns" or "variables" that we will be able to manipulate, summarize, and visualize. 





## The tidyverse

Almost everything that is useful in modern R, is contained within the `tidyverse` library, which is actually a compilation of several libraries (including `dplyr` and `ggplot2`).

```{r}
# load tidyverse library

```









## Piping

First, the most useful tool in modern R universe is the pipe: `%>%`. You can read the pipe as "**and then**". The idea of the pipe is to take the *object* on the left-hand-side of the pipe operator and place it in the first "argument" of the *function* on the right-hand-side of the pipe operator. 

This means that `object %>% function()` is equivalent to `function(object)`.

For instance, we could use piping with the `head()` function as follows:

```{r}
# compute head(gapminder) using the pipe

```


The output is itself a new object, that can then be piped into a new function, allowing us to **chain** together functions.

For instance:

```{r}
# compute the dimension of head(gapminder)

```

is equivalent to

```{r}
dim(head(gapminder))
```


but the piped version is much easier to read, since you can look at it and read *"take the `gapminder` data-frame, and then apply the `head()` function, and then apply the `dim()` function to the output"*.












## Manipulating data with dplyr functions


The functions `select()`, `mutate()`, `filter()`, `summarise()`, and `group_by()` are the workhorses behind modern data exploration and manipulation in R.


One you start using tidyverse functions, you can think of your data as the *universe*, and the columns of your data (such as `country`, `year`, and `pop`) as the *objects* that you can access within your universe. This essentially means that, within tidyverse functions, you don't need to surround the variable names in quotes.





### Select

`select()` lets you pull columns from the data. For instance the following code selects the `country` variable from the `gapminder` data frame, and then just prints out the first 6 rows. 

```{r}
# select country from gapminder and take the head

```

Note that this code just prints out the results. If we wanted to *save* the results in an object, we would need to define it as an object using the "assignment" operator `<-`, e.g.,

```{r}
# define object called top_six_countries

```

Notice how nothing was printed this time. If we want to access the results of our computation, we can write the name of the object that we defined:

```{r}
top_six_countries
```

Note that if you start thinking in the tidyverse way, you will find that you rarely actually need to define new objects. This is what makes piping so powerful: you can print out all of your results and make all of your figures without having to keep track of several different data objects.









### Filter

The `filter()` function only keeps the rows in the data that satisfy a certain condition. For instance, if you wanted to look only at the data from Australia, you can filter to rows that satisfy the condition `country == "Australia"`

```{r}
# filter to Australia

```


You can also use multiple filter conditions at the same time:

```{r}
# filter to Australia and year > 1980

```








### Mutate

The `mutate()` function lets you *add* or *modify* columns of your data. For instance, we can create a `gdp` column, by multiplying the `gdpPercap` column by the `pop` column:

```{r}
# mutate to create gdp and head

```


### Summarize

The `summarize()` function, allows you to create a summary of any column(s) you like. For instance, to compute the *mean* life expectancy:

```{r}
# compute the mean life expectency

```

The output here is a single-column single-row data frame. You can also give the summary a column name by naming your summary function:

```{r}
# compute mean life expectancy and name column

```

This is useful if you want to use the summary column you created in a downstream analysis (e.g. a visualization).










### Group by

The last function we will introduce for manipulating your data is the `group_by()` function. This allows you to use `summarize()` or `mutate()`, etc, *separately* for different values in a categorical variable.

You can think of `group_by()` as creating a separate data frame for each different level of the grouping variable, and applying whatever functions you're applying to each data frame separately, and then aggregating the results together.

For instance, to compute the average `lifeExp` separately for each `country`, you can first group by country and then apply `summarize()`:

```{r}
# compute life expectancy for each country separately

```



You should generally remember to `ungroup()` after you `group_by()` because otherwise any downstream functions you apply will still be grouped.


## Visualization with ggplot2

Let's move onto some more fun stuff: making data visualizations. The primary data visualization package in R is ggplot2, which is loaded when you load the tidyverse.

Every ggplot starts with the `ggplot()` function, which just initiates the ggplot object. 


```{r}
# use ggplot function
gapminder %>%
```



### Geom layers

To actually visualize your data, however, you need to *add* plot layers, such as a scatterplot layer using `geom_point()` or a histogram layer using `geom_histogram()`.

It's a little confusing, but when you add ggplot layers, you no longer use the pipe, but you need to use `+`:

```{r}
# add a point layer and then color by continent
gapminder %>%
  ggplot()
```














### Combining data manipulations and ggplot2

We can combine the stuff we learned above with our visualization. This plot is including *all* years of the data, if we want to just include the data from 2007, we could *filter* our data before plotting it:

```{r}
# filter to year 2007 and then do some plotting
gapminder %>%
  
```












#### Exercise: create a line plot of global average life expectency over time

```{r}
gapminder %>% 
```















### Other geoms

There are many other geom layers that you could add, e.g. `geom_line()` will create a line plot:

```{r}
gapminder %>%
  ggplot() +
  geom_line()
```



